---
title: "Retail Data Project"
author: "Nigel Barreras"
date: "2025-06-11"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1.What are the top 3 product categories by total sales?

2.Which store location had the highest average transaction amount?

3.What is the distribution of payment methods used by customers?

4.Are there any outliers in transaction amounts? If so, how would you handle them?


Server Setup
```{r server, include=TRUE}
library(DBI)
library(RSQLite)

# Create a file based SQLite database
con <- dbConnect(RSQLite::SQLite(), "retail_data.sqlite")

#read retail sales data into Rstudio enviroment
retail_data <- read.csv("retail_sales_data.csv")

# Write to SQL database
dbWriteTable(con, "retail_sales", retail_data, overwrite = TRUE)
```

CREATE TABLE category_total_sales AS 
SELECT ProductCategory, 
       SUM(TotalAmount) AS total_sales
FROM retail_sales
GROUP BY  ProductCategory
ORDER BY total_sales DESC LIMIT 3;

```{r Q1, include=TRUE}
library(DBI)
library(RSQLite)
library(ggplot2)

#Read the created sql table into the Rstudio enviroment
category_sales <- dbReadTable(con, "category_total_sales")

ggplot(data = category_sales, aes(x = reorder(ProductCategory, -total_sales),
                                  y = total_sales)) +
  geom_segment(aes(xend = ProductCategory, yend = 0), color = "gray") +
  geom_point(size = 5, color = "steelblue") +
  ylim(0,13500) +
  scale_y_continuous(breaks = seq(0, max(category_sales$total_sales),
                                  by = 2000)) +
  labs(title = "Top 3 Product Categories by Sales", x = "Product Category",
       y = "Total Sales ($)") +
  theme_minimal()

```

CREATE TABLE store_transactions AS
SELECT 
    StoreLocation,
    ROUND(AVG(TotalAmount), 2) AS avg_transaction
FROM retail_sales
GROUP BY StoreLocation
ORDER BY avg_transaction DESC;

```{r Q2, include=TRUE}

store_avg_transactions <- dbReadTable(con, "store_transactions")

ggplot(data = store_avg_transactions, 
  aes(x = reorder(StoreLocation, -avg_transaction), y = avg_transaction)) +
  geom_bar(stat = "identity", width = 0.6) +
  ylim(0,350) +
  scale_y_continuous(breaks = seq(0,     
                                  max(store_avg_transactions$avg_transaction),
                                  by = 50)) +
  labs(title = "Average Transaction amount by Store Location", x = "Location",
       y = "Transaction Amount ($)") +
  theme_minimal()

```

CREATE TABLE payment_method_dist AS
SELECT
    StoreLocation,
    PaymentMethod,
    COUNT(*) AS total_customers
FROM retail_sales
GROUP BY StoreLocation, PaymentMethod; 


```{r Q3, include=TRUE}

payment_method_dist <- dbReadTable(con, "payment_method_dist")

ggplot(data = payment_method_dist, 
  aes(x = PaymentMethod, y = StoreLocation, fill = total_customers)) +
  geom_tile(color = "grey") +
   scale_fill_gradient(low = "#CCCCCC", high = "#333333", 
                       breaks = c(6, 11, 16)) +
  labs(title = "Distribution of Payment Methods Used by Customers", 
       x = "Payment Method",
       y = "Store Location",
       fill = "Customers") +
  theme_minimal()

```

CREATE TABLE each_category_sale AS
SELECT ProductCategory,
       TotalAmount  
FROM retail_sales
WHERE TotalAmount IS NOT NULL
ORDER BY TotalAmount DESC;

```{r Q4, include=TRUE}

each_category_sale <- dbReadTable(con, "each_category_sale")

ggplot(data = each_category_sale, 
  aes(x = ProductCategory, y = TotalAmount, fill = ProductCategory)) +
  geom_boxplot(outliers = TRUE, outlier.shape = 19, outlier.colour = "red",
               show.legend = FALSE) +
  scale_y_continuous(breaks = seq(0,     
                                  max(each_category_sale$TotalAmount),
                                  by = 50)) +
  labs(title = "Distribution of Sales per Product Category", 
       x = "",
       y = "Total Sale ($)",
       fill = "Product Category",
       xlabel = "") +
  theme_minimal()

```

